---
title: "Useful Things to do with R's **Inf** Keyword"
output:
  html_document: default
  html_notebook: default
---
```{r, include=FALSE}
require(ggplot2)
```

Having used R for some years now, I've come across a few instances where R's **Inf** keyword can be used to yield a practical outcome. For those unfamiliar with R's **Inf** keyword, it is defined as a postive or negative number divided by zero yielding postive or negative **Inf**, respectively.

```{r definition of Inf}
c(plus_inf = 1/0, minus_inf = -1/0)

```

So how we can make practical use of infinity in R? Here are a few applications I've found: 

  1. *cut()* defining the upper and lower extremes.
  1. *ggplot2::annotate()* positioning text in ggplot at a corner or edge.
  1. *integrate()* when setting the upper and lower integral bounds

#**Inf** with *cut()*# 

Suppose you want to bin the following set of numbers into five discrete levels.  

<pre>{-100, -1,1,1,1,2,2,2,5,5,5,10,10,10,100,1000}</pre>  

You might do something like, 

```{r using the cut() Function}
numbers <- c(-100, -1,1,1,1,2,2,2,5,5,5,10,10,10,100,1000)
bins <- cut(numbers, breaks = c(-101,-2,2,5,11, 1001))
xtabs(~bins)
```

The above code explicitly shows how we might bin the data including the extreme values -100,100, and 1000. But with real data, extrama can change when new data is added. For example, what if we received the following new pieces of data: -102 and 1002? Below, we've added them to our list and re-binned them as before.  

```{r using the Cut Function Not Robust}
numbers <- c(-102, -100, -1,1,1,1,2,2,2,5,5,5,10,10,10,100,1000,1002)
bins <- cut(numbers, breaks = c(-101,-2,2,5,11, 1001))
xtabs(~bins)
```

Opps! The new data was not included in the binning. We would need to manually expand our limits manually, or we could perhaps do something clever with the *max* and *min* functions. 

Alternativly, we can solve this problem with R's $\pm$ **Inf** keyword by placing $\pm$ **Inf** in the *breaks* argument of the *cut* function (*see line 2 below*)


```{r using the Cut Function with Infs}
numbers <- c(-102, -100, -1,1,1,1,2,2,2,5,5,5,10,10,10,100,1000,1002)
bins <- cut(numbers, breaks = c(-Inf,-2,2,5,11, Inf))
xtabs(~bins)
```

Now, when new data is collected that is less than -2 or greater than 11 there will be a bin to catch it, no mater how big or how small. 

#**Inf** in integrate()#

Integration is never fun, but sometimes its necessary. For example, to get the probability of a normally distributed event between two limits you would need to integrate the normal distribution density funciton between limits a and b. 

$$
f(x |\mu, \sigma^2) =  \frac{1}{\sqrt{2\pi\sigma^2}} e^{-\frac{(x - \mu)^2}{2\sigma^2}} \qquad \mathrm{Density \space Function \space(dnorm)} 
\\
P(x |\mu, \sigma^2) = \int\limits_a^b \frac{1}{\sqrt{2\pi\sigma^2}} e^{-\frac{(x - \mu)^2}{2\sigma^2}} \qquad \mathrm{Proability \space Function \space (pnorm)} 

$$

The total area under the Density Function should be 1. So if we integrate dnorm(x) from negative to positive infinity we should get 1. Let's see

```{r}
LongDF <- function(x, mu = 0 , sigma = 1){
            1/sqrt(2*pi*sigma^2) * exp(-(x-mu)^2/(2*sigma^2))
          }

#dnorm and the long function are the same...
#dnorm(0) == LongDF(0)
#TRUE

integrate(dnorm, -Inf,  Inf)
```

as expected the answer is 1. 


#**Inf** in ggplot2::annotate() - text positioning#  

In the documentation for ggplot2 http://docs.ggplot2.org/current/geom_text.html <end of page> you are told you can position text in a plot in the following way.   

```{r as shown by ggplot documentation}
df <- data.frame(
  x = c(1, 1, 2, 2),
  y = c(1, 2, 1, 2),
  text = c("bottom-left", "bottom-right", "top-left", "top-right")
)

ggplot(df, aes(x, y)) + geom_text(aes(label = text), vjust = "inward", hjust = "inward") 
```

Ok, this puts the text in the corners, but look what happens when we change the plot limits with *ylim()* and *xlim()*. 

```{r not at the corners anymore}
ggplot(df, aes(x, y)) + geom_text(aes(label = text), vjust = "inward", hjust = "inward") + 
  ylim(-10,10) + xlim(-10,10)
```

Oops. The text is no longer in the corners. Using R's **Inf** keyword, we can achor the text in a simple, robust way to the corners and edges of a plot. Here is the plot again with original limits above using **Inf**. I've also placed some extra annotations in the middle of each side for demonstration.  

```{r Using Inf to lock text at edges}
gg_corner_edge_text <- ggplot(df, aes(x,y)) +  
  annotate("text", label="Top Left",  x = -Inf, y=Inf, hjust=0, vjust=1) +
  annotate("text", label="Bottom Left",  x = -Inf, y=-Inf, hjust=0, vjust=0) +
  annotate("text", label="Top Right",  x = Inf, y=Inf, hjust=1, vjust=1) +
   annotate("text", label="Bottom Right",  x = Inf, y=-Inf, hjust=1.1, vjust=-.7) +

  #### Placeing text at Top or Bottom Givben X
  annotate("text", label="Top Middle",  x=(max(df$x)+min(df$x))/2 , y=Inf, hjust=.5, vjust=1) +
  annotate("text", label="Bottom Middle",  x=(max(df$x)+min(df$x))/2 , y=-Inf, hjust=.5, vjust=-.7) +

  #### Placeing text at Left or Right Side Given Y
  annotate("text", label="Right Middle", x=Inf , y=(max(df$y)+min(df$y))/2, hjust=.5, vjust=1, angle=-90) +
  annotate("text", label="Left Middle",  x=-Inf , y=(max(df$y)+min(df$y))/2, hjust=.5, vjust=1, angle=90)

gg_corner_edge_text
```

Here is the same plot, except now we've expanded the limits with *xlim()* and *ylim()* as above to show that the text stays on the edges. 


```{r SomeData}
gg_corner_edge_text + ylim(-10,10) + xlim(-10,10)
```

Awesome! The text stayed in the corners, and center edge of each side. However, it is no longer centered correctly. At this point, I do not have a robust way to deal with centering, but the text is locked to the edges of the plot because of the **Inf** keyword. Fine adjustment to the alignment can be achieved with hjust and vjust agruments.


#Summary

In this brief post, we shown how to use R's **Inf** keyword to bin data using the *cut()* function and position text at the corners and edges of plots made with ggplot. 


